Modal Abstractions for Virtualizing Memory Addresses

### 2024-06-28 examples.tex
compiles

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline|pte\_walkpgdir (discussed again in Section \ref{sec:traversing}).
\lstinline|pte\_walkpgdir|'s postcondition contains physical pointsto assertions for a partial page table walk reaching the
L1 entry (l1e), ensuring higher levels of the page table exist. 
The postcondition also includes a virtual \emph{PTE}
pointsto ($\mapsto_{\textsf{vpte}}$) allows access to the memory of the L1 entry
via virtual address \lstinline|pte_pointsto|.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte}} \;\{\textsf{q}\} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\sumapaces{\mathit{cr3val}}\delta) \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
This is also guaranteed ($ \ulcorner
\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.\todo{should we also discuss the following before while introducing the invariant?} The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
\subsubsection{Traversing Page Tables}

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{OFF} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, OFF ;; shift physical address by identity mapping constant OFF
$\specline{\textsf{P} \ast \exists \qfraczero; \qfrac,\textsf{ w, val},  \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{OFF} \ast \sumapaces\rtv\delta \ast \sumpv\paddr\qfrac\true \ast \vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
$\specline{\ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr \ast (R \wand \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)) }_{\rtv}$
\end{lstlisting}
\vspace{-1em}
\caption{Traversing page-tables, and allocating entries as needed while mapping-a-page in Figure \ref{fig:mappingcode}.}
\label{fig:walkpgdir}
\end{figure}

>>>>

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline|pte\_walkpgdir| (discussed again in Section \ref{sec:traversing}).
\lstinline|pte\_walkpgdir|'s postcondition contains physical pointsto assertions for a partial page table walk reaching the
L1 entry (l1e), ensuring higher levels of the page table exist. 
The postcondition also includes a virtual \emph{PTE}
pointsto ($\mapsto_{\textsf{vpte}}$) allows access to the memory of the L1 entry
via virtual address \lstinline|pte_pointsto|.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte}} \;\{\textsf{q}\} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\sumapaces{\mathit{cr3val}}\delta) \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
This is also guaranteed ($ \ulcorner
\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
\subsubsection{Traversing Page Tables}

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
push   rbp
mov    rsp,rbp
sub    0x20,rsp
mov    rdi,-0x8[rbp]

mov    -0x8[rbp],rdi
mov    [rdi],rdi
and    0x1,rdi
mov    rdi,rax
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>

lea    -0x10[rbp],rdi
callq  70 <pte_initialize>
mov    -0x8[rbp],rdi
mov    [rdi],rax
and    0xfffffffffffffffd,rax
or     0x2,rax
mov    rax,[rdi]

mov    -0x8[rbp],rax
mov    [rax],rdi
and    0xfffffffffffffffb,%rdi
or     0x4,rdi
mov    rdi,[rax]

mov    -0x10[rbp],rax
movabs 0x0,rdi	115: R_X86_64_64	KERN_BASE
sub    rdi,rax
shr    0xc,rax
mov    -0x8[rbp],rdi
mov    [rdi],rcx
movabs 0xffffffffff,rdx
and    rdx,rax
shl    0xc,rax
movabs 0xfff0000000000fff,rdx
and    rdx,rcx
or     rax,rcx
mov    rcx,[rdi]

mov    -0x8[rbp],rax
mov    [rax],rcx
and    0xfffffffffffffffe,rcx
or     0x1,rcx
mov    rcx,[rax]

mov    -0x8[rbp],rax
mov    [rax],rax
shr    0xc,rax
movabs 0xffffffffff,rcx
and    rcx,rax
shl    0xc,rax
mov    rax,-0x18[rbp]

mov    -0x18[rbp],rax
movabs 0x0,rcx	187: R_X86_64_64	KERN_BASE
add    rcx,rax
mov    rax,-0x20[rbp]
mov    -0x20[rbp],rax
mov    rax,-0x10[rbp]

 mov    -0x10[rbp],rax
 add    0x20,rsp
 pop    rbp
 ret
\end{lstlisting}
\vspace{-1em}
\caption{Traversing page-tables, and allocating entries as needed while mapping-a-page in Figure \ref{fig:mappingcode}.}
\label{fig:walkpgdir}
\end{figure}
